---
title: Matrices
subtitle: Introduction
toc-title: Matrices--Introduction
---

```{r}
#| label: loadermatrices
#| include: false
#| cache: false
source("loader.R")
```

A matrix is a series of numbers arranged in rows and columns. We usually abbreviate matrices with a bolded uppercase letter:


$$
\mathbf{A}_{j \times k} =
\begin{bmatrix}
  a_{1,1} & a_{1,2} & \cdots & a_{1,k} \\
  a_{2,1} & a_{2,2} & \cdots & a_{2,k} \\
  \vdots  & \vdots  & \ddots & \vdots  \\
  a_{j,1} & a_{j,2} & \cdots & a_{j,k}
\end{bmatrix}
$$


For example, 

$$
\mathbf{A} =
\begin{bmatrix}
  1&2&3&4\\
  5&6&7&8\\
  9&10&11&12
\end{bmatrix}
$$

To make a matrix in R, there are many possibilities. For the sake of simplicity, I will recommend using the `matrix` function with the `byrow` option set to `TRUE`. 

The `matrix` function takes a vector and converts it to a matrix. By default, the `matrix` function assumes that the vector is a single column. For example,

```{r demo-matrix1column}
matrix(1:4)
```

By setting the `nrow` (number of rows) or `ncol` (number of columns), the matrix dimensions are set. For example,

```{r demo-ncol}
matrix(1:4, ncol = 2)
```

Note that the default is to fill in the matrix by columns. If you want it to fill by rows, set the `byrow` option to `TRUE`.

```{r demo-byrow2}
matrix(1:4, ncol = 2, byrow = TRUE)
```

In general, creating matrices by row makes the code and the matrix look similar:

```{r demo-byrowsimilar}
matrix(c(3, 5, 
         2, 3),
       nrow = 2,
       byrow = TRUE)
```

To achieve the same matrix without the `byrow` option set to `TRUE` is possible, but the code will be the same order as the matrix:

```{r demo-bycolnotsimilar}
matrix(c(3, 2, 
         5, 3),
       nrow = 2)
```



::: {.callout-note title="You Try"}

Make vector `A` into a 2 by 2 matrix like so:

$$
\mathbf{A}=\begin{bmatrix}
  2&3\\
  5&7
 \end{bmatrix}
$$

```{webr-r}
#| label: ex-makeA
A <- matrix(c(_,_,
              _,_), 
            nrow = _, 
            byrow = _)

```


```{r solution-binding}

# Using the matrix function setting byrow = TRUE
A <- matrix(c(2, 3,
              5, 7),
            nrow = 2,
            byrow = TRUE)

```

:::


## Appending Matrices

$$
\color{RoyalBlue}{\mathbf{A}}=\begin{bmatrix}
\color{RoyalBlue}{1} & \color{RoyalBlue}{2} & \color{RoyalBlue}{3} & \color{RoyalBlue}{4} \\ 
\color{RoyalBlue}{5} & \color{RoyalBlue}{6} & \color{RoyalBlue}{7} & \color{RoyalBlue}{8} \\ 
\color{RoyalBlue}{9} & \color{RoyalBlue}{10} & \color{RoyalBlue}{11} & \color{RoyalBlue}{12}
\end{bmatrix}, \color{Firebrick}{\mathbf{B}}=\begin{bmatrix}
\color{Firebrick}{13} & \color{Firebrick}{14} & \color{Firebrick}{15} & \color{Firebrick}{16} \\ 
\color{Firebrick}{17} & \color{Firebrick}{18} & \color{Firebrick}{19} & \color{Firebrick}{20}
\end{bmatrix}
$$

The equation below means, "Make a new matrix called C by appending B to the bottom of A."

$$
\mathbf{C}=\begin{bmatrix}
\color{RoyalBlue}{\mathbf{A}} \\ \color{FireBrick}{\mathbf{B}} 
\end{bmatrix}=
\begin{bmatrix}
\color{RoyalBlue}{1} & \color{RoyalBlue}{2} & \color{RoyalBlue}{3} & \color{RoyalBlue}{4} \\ 
\color{RoyalBlue}{5} & \color{RoyalBlue}{6} & \color{RoyalBlue}{7} & \color{RoyalBlue}{8} \\ 
\color{RoyalBlue}{9} & \color{RoyalBlue}{10} & \color{RoyalBlue}{11} & \color{RoyalBlue}{12}\\
\color{Firebrick}{13} & \color{Firebrick}{14} & \color{Firebrick}{15} & \color{Firebrick}{16} \\ 
\color{Firebrick}{17} & \color{Firebrick}{18} & \color{Firebrick}{19} & \color{Firebrick}{20}
\end{bmatrix}
$$

In R, adding new rows to a matrix is done with the `rbind` function (Think "row bind"). To append rows, matrices must have the same number of columns to be compatible.

```{r demo-appendrbind}
A <- matrix(1:12, nrow = 3, byrow = TRUE)
B <- matrix(13:20, nrow = 2, byrow = TRUE)
C <- rbind(A,B)
C
```

The `cbind` function (Think "column bind") works the same way but it appends columns to the right of a matrix. Matrices must have the same number of rows to be compatible.

```{r demo-appendcbind}
A <- matrix(1:4, nrow = 2)
B <- matrix(5:8, nrow = 2)
C <- cbind(A,B)
C
```

::: {.callout-note title="You Try"}

Make a 2 by 2 matrix **A**:

$$
A=\begin{bmatrix}
  11&13\\
  17&19
\end{bmatrix}
$$

Make a 2 by 2 matrix `B`:

$$
B=\begin{bmatrix}
  11&13\\
  17&19
\end{bmatrix}
$$


Append `A` and `B` with `rbind` to make a 4 by 2 matrix like so:

$$
\begin{bmatrix}
  2&3\\
  5&7\\
  11&13\\
  17&19
\end{bmatrix}
$$


```{webr-r}
#| label: ex-rbinding
A <- matrix(c(_,_,
              _,_), 
            nrow = _, 
            byrow = _)
B <- matrix(c(_,_,
              _,_), 
            nrow = _, 
            byrow = _)

# Now bind A and B by rows
rbind(_,_)
```

```{r solution-rbindAB}
A <- matrix(c(2, 5, 3, 7), nrow = 2)
B <- matrix(c(11,17,13,19), nrow = 2)
rbind(A, B)
```

Now use `cbind` to make a 2 by 4 matrix:

$$
\begin{bmatrix}
  2&3&11&13\\
  5&7&17&19
\end{bmatrix}
$$


```{webr-r}
#| label: ex-cbinding
# Now bind A and B by columns
cbind(_,_)
```

```{r solution-cbindAB}
cbind(A, B)
```

:::



## Selecting parts of a matrix 

Let's make this matrix:

$$
\mathbf{A} =
\begin{bmatrix}
  1&2&3&4\\
  5&6&7&8\\
  9&10&11&12
\end{bmatrix}
$$

```{r demo-byrow}
A <- matrix(1:12, nrow = 3, byrow = TRUE)
A
```


### Select a single element of a matrix

To select a single element of a matrix, specify the row and column in brackets after the matrix. For example, the element $\mathbf{A}_{3,2}$ (i.e., the 3^rd^ row and 2^nd^ column of $\mathbf{A}$) is 

```{r demo-select32}
A[3,2]
```

::: {.callout-note title="You Try"}

Select the element in row 2, column 3

```{webr-r}
#| label: ex-select
A <- matrix(1:12, nrow = 3, byrow = TRUE)
A[_,_]
```

```{r solution-select}
A[2,3]
```

:::

### Select a matrix row

Leaving one of the slots in the bracket empty means that you want all of the elements in that row or column. 

$\mathbf{A}_{1\bullet}$ is the 1^st^ row of $\mathbf{A}$.

```{r demo-selectrow}
A[1, ]
```

### Select a matrix column

$\mathbf{A}_{\bullet 3}$ is the 3^rd^ column of $\mathbf{A}$.

```{r demo-selectcolumn}
A[, 3]
```

By default, whenever a single row, column, or element is returned from a matrix, R drops the row and column dimensions. If you wish to preserve the result in matrix form, set `drop` to `FALSE`:

```{r demo-selectdrop}
A[, 3, drop = FALSE]
```

::: {#callout-note title="You Try"}

Select column 1 of A.

```{webr-r}
#| label: ex-selectA1
A
```

```{r solution-matrixrow}
A[, 1]
```
:::



### Select several columns or rows

A vector of integers will select whichever rows or columns you wish. Here are the 2^nd^ and 3^rd^ rows:

```{r demo-select23rows}
A[2:3, ]
```

Here are the 1^st^ and 4^th^ columns:

```{r demo-select14cols}
A[, c(1, 4)]
```

::: {#callout-note title="You Try"}

Select columns 1 and 3 of A. 

```{webr-r}
#| label: ex-matrixcols
A
```

```{r solution-matrixcols}
A[, c(1, 3)]
```
:::


### Selecting with Boolean vectors

Here is the first two rows of $\mathbf{A}$:

```{r demo-selectboolean}
A[c(TRUE,TRUE,FALSE),]
```

This seems like a strange way to do this but it is actually quite powerful. Any vector of `TRUE` and `FALSE` values can be used to select things. For example, select a column only if its first value is greater than 2:

```{r demo-selectgreaterthan2}
s <- A[1,] > 2 # Creates a vector that tests whether the first row elements are greater than 2
A[,s] # Select only the columns whose first value is greater than 2
```


::: {#callout-note title="You Try"}

Use a Boolean vector to select columns that begin with 2. The operator for "is equal to" is `==`.

```{webr-r}
#| label: ex-boolean
s <- A[] == _ 
A[,s] 
```

```{r solution-boolean}
s <- A[1,] == 2 
# When only one column or row is selected, R returns a vector
A[,s]
# To force R to return a matrix, set drop = FALSE
A[,s, drop = FALSE]
```
:::


### Selecting with name vectors

We can give row and column names to a matrix like so:

```{r demo-selectnamevector}
rownames(A) <- c("Gold","Silver","Bronze")
colnames(A) <- c("Vault","Uneven Bars","Balance Beam","Floor")
A
```

Now we can select rows and columns by names:

```{r demo-selectgoldbronze}
A[c("Gold","Bronze"),]
```

::: {#callout-note title="You Try"}

Use the column names to select the Uneven Bars column

```{webr-r}
#| label: ex-uneven
A <- matrix(1:12, nrow = 3, byrow = TRUE)
rownames(A) <- c("Gold","Silver","Bronze")
colnames(A) <- c("Vault","Uneven Bars","Balance Beam","Floor")
# select the Uneven Bars column
A
```

```{r solution-uneven}
A[,"Uneven Bars"]
# or to force returning a matrix
A[,"Uneven Bars", drop = FALSE]
```

Select the intersection of Silver and Floor

```{webr-r}
#| label: ex-silverfloor
A
```

```{r solution-silverfloor}
A["Silver", "Floor"]
# or
A["Silver", "Floor", drop = FALSE]
```

:::

## Replace portions of a matrix

Any portion of a matrix can be replaced with new values. For example, this will replace the first row with zeros:

```{r demo-replacerow}
A[1, ] <- c(0, 0, 0, 0)
A
```

This can be done by column or row name as well

```{r demo-replacerowgold}
A["Gold", ] <- c(0, 0, 0, 0)
A
```


::: {#callout-note title="You Try"}

Replace the Vault column of $\mathbf{A}$ with a vector of $(10,20,30)$

```{webr-r}
#| label: ex-replacevault
A
```

```{r solution-replacevault}
A[, "Vault"] <- c(10, 20, 30)
A
```
:::

## Transposing matrices

[To *transpose* a matrix is to write its rows as columns and its columns as rows.]{.defword .column-margin}

To [transpose]{.defword} a matrix is to flip its rows into columns.

$\mathbf{A}'$ is matrix $\mathbf{A}$ transposed.

If

$$\mathbf{A}=\begin{bmatrix}
1&2&3\\
4&5&6
\end{bmatrix}$$

Then 

$$\mathbf{A}'=\begin{bmatrix}
1&4\\
2&5\\
3&6
\end{bmatrix}$$

### Transposing in R

In R the `t` function transposes matrices.

```{r demo-transpose}
A <- matrix(1:6, nrow = 2, byrow = TRUE)
A
Aprime <- t(A)
Aprime
```

::: {#callout-note title="You Try"}

1. Make a 2 $\times$ 5 matrix `A` of even numbers up to 20. 
2. Transpose `A`, assigning it to a variable called `Aprime`.
3. Replace the last row (row 5) of `Aprime` with zeroes.

```{webr-r}
#| label: ex-even20transpose
A <- matrix(seq(_,_,_), nrow = _)
```


```{r solution-even20transpose}
A <- matrix(seq(2, 20, 2), nrow = 2)
A
Aprime <- t(A)
Aprime
Aprime[5, ] <- c(0, 0)
Aprime
```
:::



## Square matrices

In a **square matrix**, the number of rows is equal to the number of columns.

## Symmetric matrices

A **symmetric matrix** is a square matrix that is equal to its transpose.

$$\mathbf{A}=\mathbf{A}'$$

This means that for all elements, $a_{ij}=a_{ji}$.

Here is an example of a symmetric matrix:

$$\begin{bmatrix}
\color{green}a & \color{Firebrick}b & \color{RoyalBlue}c\\
\color{Firebrick}b & \color{gold}d & \color{DarkOrchid}e\\
\color{RoyalBlue}c & \color{DarkOrchid}e & \color{orange}f
\end{bmatrix}$$

To verify that a matrix is symmetric in R:

```{r demo-symmetrictest}
#| eval: false
all(A == t(A))
# Or use a dedicated function
isSymmetric(A)
```

Correlation matrices and covariance matrices are always symmetric.

## Diagonal matrices

A **diagonal matrix** is a square matrix consisting of zeroes everywhere except the diagonal. For example,

$$
\mathbf{A} = \begin{bmatrix}
a & 0 & 0\\
0 & b & 0\\
0 & 0 & c
\end{bmatrix}
$$

To create a diagonal matrix, specify the diagonal vector and then insert it into the `diag` function like so:

```{r demo-MakeDiagonal}
a <- 1:4
A <- diag(a)
A
```

As we will see later, the `diag` function actually has several different purposes:

1. It creates a diagonal matrix $\mathbf{A}$ from a vector $\vec{a}$.  
    `A <- diag(a)` 
2. It extracts a diagonal vector $\vec{a}$ from a matrix $\mathbf{A}$.  
    `a <- diag(A)`
3. It creates an identity matrix $\mathbf{I}$ from a positive integer $n$.  
    `I <- diag(n)`
4. It replaces the diagonal of matrix $\mathbf{A}$ with a new vector $\vec{b}$.  
    `diag(A) <- b`

